PyPy's Approach to Virtual Machine Construction

### 2006-07-12 talk/dls2006/paper.tex
give a better feeling of the expressiveness of helpers.

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environments.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
``restricted Python'' or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough type information to generate
efficient code.  RPython still supports inheritance but limited to
single inheritance with some mix-in support, dynamic dispatch, to some
extent keywords arguments and varargs, first-class function and class
values, limited use of bound methods, runtime 'isinstance' and type
queries, but no runtime reflection, further bindings in class and
global namespaces are assumed constant. RPython code can be run on a
Python interpreter without severe semantics mismatches, figure
\ref{fig_create_frame} shows some RPython code from the
\textit{Standard Interpreter}, it is still quite idiomatic Python code
using the built-in dictionary type and first-class class values
instantiation.

@@

@@

>>>>

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environments.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
``restricted Python'' or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough type information to generate
efficient code.  RPython still supports inheritance but limited to
single inheritance with some mix-in support, dynamic dispatch, to some
extent keywords arguments and varargs, first-class function and class
values, limited use of bound methods, runtime \texttt{isinstance} and type
queries, but no runtime reflection, further bindings in class and
global namespaces are assumed constant. RPython code can be run on a
Python interpreter without severe semantics mismatches, figure
\ref{fig_create_frame} shows some RPython code from the
\textit{Standard Interpreter}, it is still quite idiomatic Python code
using the built-in dictionary type and first-class class values
instantiation.

@@

Because our helpers can be both run on top of CPython and translated
we can reuse them both as runtime implementations and to render constant
objects at translation time. This is both natural and useful (especially for complex
data structures). Figure \ref{fig_insertclean} shows a function
implementing insertion in a built-in dictionary for keys known to be
absent, this is used at runtime after resizing and to fill the constant
dictionaries at translation.

\begin{figure}
\begin{verbatim}
def ll_dict_insertclean(d, key, value, hash):
  entry = ll_dict_lookup_clean(d, hash)
  ENTRY = lltype.typeOf(entry).TO
  entry.value = value
  entry.key = key
  if hasattr(ENTRY, 'f_hash'): 
    entry.f_hash = hash
  if hasattr(ENTRY, 'f_valid'):    
    entry.f_valid = True
  if hasattr(ENTRY, 'f_everused'): 
    entry.f_everused = True
  d.num_items += 1
  d.num_pristine_entries -= 1
\end{verbatim}
\label{fig_insertclean}
\caption{built-in dict insertion helper.}
\end{figure}

@@

It is also simple, in the context of abstract interpretation, to
perform translation-time constant propagation, computations and
implement conditional code in a natural way. For example, the code in
figure \ref{fig_insertclean}, which is then specialized for the
possible types of dictionary \texttt{d}, shows this kind of
compile-time computation and introspection on low-level types,
\texttt{ENTRY} is a C-like structure type and the
\texttt{hasattr}-ibute checks introspect it at translation time. This
kind of expressiveness increased our ability of reuse code even in
low-level helpers.
